---
title: "Database Acquisition"
author: "Madie Munro"
date: "4/6/2023"
output:
  word_document: default
  html_document: default
---

# R Setup:

**In order to set the correct project directory for this stage of development, go to:**

*Tools > Global Options > R Markdown*

**Once there, set "Evaluate chunks in directory" from "Document" to "Project." Then, select "Apply" and hit "OK" when done.**

**The code chunk below imports libraries needed for R setup including knitr to knit the file to Word or HTML and here to read in the utility R file used for certain functions for data wrangling. It will not display when knitted to Word or HTML.**

```{r setup, include=FALSE, error=FALSE}
library(knitr)
library(here)
opts_chunk$set(echo=T)
# knitr::opts_knit$set(root.dir = '.')
source(here("01_acquisition/02_protocol/split_utils.R"))
```

# Introduction:

## File Reading:

**The following code chunk below imports the needed libraries for table ripping: **

```{r, error=FALSE, warning=FALSE, results='hide'}
library(tidyverse)
library(dplyr)
library(naniar)
library(stringr)
```

**Next, the original dataset was converted from an Excel file to a CSV file in order to read in the data easier (see 03_incremental folder for the .csv version of the data). In addition, a column lookup dataset was read in, containing the new names for the columns of the database as well as a description of the type of data stored in each respective column. Lastly, extra na values were dropped after the table was read in: **

```{r, error=FALSE, warning=FALSE, results='hide'}
fish <- 
  read_csv("./01_acquisition/03_incremental/fish_info.csv")

fish_cols <- 
  read_csv("./01_acquisition/01_input/column_lookup.csv")  
fish_cols <- drop_na(fish_cols)
```

## Column Renaming: 

**The column names stored in "fish" were changed to reflect the new column names to be used in the updated database. This was accomplished by creating a subset of the "fish_cols" data, pivoting the data wider in order for the new column names to be the dataframe column names, and then renaming each column name in "fish" to the column names in the subsetted data. Lastly, the column "FishIDX" was added to the end of the "fish" dataset to serve as an indexing variable for each data entry: **

```{r, error=FALSE, warning=FALSE}
cols_sub <- fish_cols[, c("NewName", "ColDesc")]
fcols <- pivot_wider(cols_sub, names_from = "NewName", values_from = "ColDesc")

colnames(fish) = colnames(fcols)
fish$FishIDX <- 1:nrow(fish)
```

## Data Mutation: 

**Some numeric column values gave a range of values rather than one set value, so they were split into two respective columns: low and high. The first column to be split as such was the water temperature column. This was achieved by the utility file, "split_utils.R, " which contains two functions: one for splitting specifically on the plus-minus character and one general character splitting function (see 01_acqusition/02_protocol/split_utils.R for code):**

```{r, warning=FALSE, results='hide'}

#creates a variable with the old temp measurements
fish_temps <-
  fish$Temperature %>%
  as.list() 

#correctly splits up values that are ranges or plus-minus values (ex: 15 plus-minus 3, 12 to 19)
fish_temps_to_replace <-
  pm_split(fish_temps) 
fish_temps_to_replace <-
  c(fish_temps_to_replace,
  char_split(fish_temps, "to"))

lapply(
  fish_temps_to_replace,
  function(idx){
    item_in_fish_temps_list <- idx[[1]]
    fish_temps[[item_in_fish_temps_list]] <<- idx[[2]]
  }
)

#turns list elements into two separate dataframe columns
fish_temps <- lapply(fish_temps, function(x) { 
  x <- 
    as.vector(x)%>%
    as.numeric 
  if(length(x) == 2){
    return(x)
  }
  else {
    x <- rep(x,2)
  }
}) %>% do.call(rbind, .) %>% 
  as.data.frame(.)

#creates low and high temp columns

fish_temps_low <- fish_temps$V1
fish_temps_high <- fish_temps$V2

```

**The swim variability column also got the same treatment: **

```{r, warning=FALSE, results='hide'}
swim_var_temp <-
  fish$SwimVariabilityValue %>%
  as.list()

swim_var_to_replace <-
  char_split(swim_var_temp)

lapply(
  swim_var_to_replace,
  function(idx){
    item_in_swim_var_list <- idx[[1]]
    swim_var_temp[[item_in_swim_var_list]] <<- idx[[2]]
  }
)

swim_var_temp <- lapply(swim_var_temp, function(x) { 
  x <- 
    as.vector(x)%>%
    as.numeric 
  if(length(x) == 2){
    return(x)
  }
  else {
    x <- rep(x,2)
  }
}) %>% do.call(rbind, .) %>% 
  as.data.frame(.)

fish_swim_var_low <- swim_var_temp$V1
fish_swim_var_high <-swim_var_temp$V2
```

**Once the more problematic columns were appropriately split apart, each of the columns with numeric values were converted from character to numeric types. Some values were converted to double, some were converted to integer. The modified data was stored under "fish_test": **

```{r, warning=FALSE}
fish_test <- fish %>%
  mutate_at(c("MinSpeed", "MaxSpeed", "SwimSpeed", "SwimVariabilityValue", "MinLength", "MaxLength", "LengthValue", "LengthVariabilityValue",
              "SwimTime", "MinTime", "MaxTime", "TimeBetweenIncrements", "VelocityIncrements", "SwimmingTimeAtMaxVelocity"), as.numeric) %>%
  mutate(NumberOfFishTested = as.integer(NumberOfFishTested)) %>%
  mutate(fish_temps_high = as.double(fish_temps_high)) %>%
  mutate(fish_temps_low = as.double(fish_temps_low)) %>%
  mutate(fish_swim_var_high = as.double(fish_swim_var_high)) %>%
  mutate(fish_swim_var_low = as.double(fish_swim_var_low))

```

**The last steps taken before the full data was ripped into separate tables for SQL querying was the renaming of the new columns (low/high temp/swim variability) and replacing the question marks in the "Leap(y/n)" column with "na": **

```{r, warning=FALSE}
colnames(fish_test)[42] <- "WaterTempHigh" 
colnames(fish_test)[43] <- "WaterTempLow" 
colnames(fish_test)[44] <- "SwimVariabilityValueHigh"
colnames(fish_test)[45] <- "SwimVariabilityValueLow"

fish_test <- fish_test %>% replace_with_na_at(.vars = "Leap(Y/N)", condition = ~.x == "?")

```

# Individual Table Creation: 

## Subsetting Original Data:

**The first step taken to create the individual database tables was to create multiple subsets of the wrangled data based on the database relational model drafted (see 03_incremental for model): **

```{r, warning=FALSE}
fish_info <- fish_test[, c("FishIDX", "CommonName", "ScientificName")]

swim_info <- fish_test[, c("SwimMode", "SwimSpeed", "SwimUnits", "SwimTime", "Leap(Y/N)")]

fish_len_info <- fish_test[, c("LengthValue", "LengthUnits", "LengthType", "BD/TLRatio", "TL/SLRatio", "TL/FLRatio")]

exp_method <- fish_test[, c("AveragingCoeff(Y/N)", "NumberOfFishTested", "TestMethod", "Reference", "DataReference", "Comments")]

equation <- fish_test[, c("VelocityEquationGiven", "BurstCoeffA", "BurstCoeffB", "BurstCoeffC", "ProlongedCoeffA", "ProlongedCoeffB", "ProlongedCoeffC")]

water_temp <- fish_test[, c("WaterTempLow", "WaterTempHigh")]

swim_ex_info <- fish_test[, c("MinTime", "MaxTime", "TimeBetweenIncrements", "SwimmingTimeAtMaxVelocity", "VelocityIncrements")]

swim_var <- fish_test[, c("MinSpeed", "MaxSpeed", "SwimUnits", "SwimVariabilityValueLow", "SwimVariabilityValueHigh", "SwimVariabilityType")]

fish_len_var <- fish_test[, c("MinLength", "MaxLength", "LengthUnits", "LengthVariabilityValue", "LengthVariabilityType")]
```

## Adding/Rearranging Relational Model Columns: 

**Since the database relational model features columns in the tables not mentioned in the original or wrangled database (ex: there aren't any unique ID values), they had to be created from scratch. This was done by creating two separate columns, one numeric and one string, and then uniting them together. Additional units for certain measurements like swim velocity, fish length, and temperature were also added to tables in addition to a column for the central tendency metric used for numeric measurements: **

```{r, warning=FALSE}
#Fish Information
fish_info <- fish_info %>%
  mutate(FishIDNum = row_number())
fish_info <- fish_info %>%
  mutate(FishIDStr = "FSH")
fish_info <- fish_info %>%
  unite(FishID, FishIDNum, FishIDStr, sep="")

fish_info <- fish_info %>%
  mutate(SwimIDNum = row_number())
fish_info <- fish_info %>%
  mutate(SwimIDStr = "SWM")
fish_info <- fish_info %>%
  unite(SwimID, SwimIDNum, SwimIDStr, sep="")

fish_info <- fish_info %>%
  mutate(ExpMethodIDNum = row_number())
fish_info <- fish_info %>%
  mutate(ExpMethodIDStr = "EXP")
fish_info <- fish_info %>%
  unite(ExpMethodID, ExpMethodIDNum, ExpMethodIDStr, sep="")

#Swimming Information
swim_info <- swim_info %>%
  mutate(SwimIDNum = row_number())
swim_info <- swim_info %>%
  mutate(SwimIDStr = "SWM")
swim_info <- swim_info %>%
  unite(SwimID, SwimIDNum, SwimIDStr, sep="")

swim_info <- swim_info %>%
  mutate(SwimExIDNum = row_number())
swim_info <- swim_info %>%
  mutate(SwimExIDStr = "EXH")
swim_info <- swim_info %>%
  unite(SwimExID, SwimExIDNum, SwimExIDStr, sep="")

swim_info <- swim_info %>%
  mutate(WaterTempIDNum = row_number())
swim_info <- swim_info %>%
  mutate(WaterTempIDStr = "WTP")
swim_info <- swim_info %>%
  unite(WaterTempID, WaterTempIDNum, WaterTempIDStr, sep="")

swim_info <- swim_info %>%
  mutate(SwimVarIDNum = row_number())
swim_info <- swim_info %>%
  mutate(SwimVarIDStr = "SVR")
swim_info <- swim_info %>%
  unite(SwimVarID, SwimVarIDNum, SwimVarIDStr, sep="")

swim_info <- swim_info %>%
  mutate(EqIDNum = row_number())
swim_info <- swim_info %>%
  mutate(EqIDStr = "EQ")
swim_info <- swim_info %>%
  unite(EqID, EqIDNum, EqIDStr, sep="")

swim_info <- swim_info %>%
  mutate(SwimCentrTendMetric = "mean")

swim_info <- swim_info %>%
  mutate(TimeUnits = "s")

#Equations
equation <- equation %>%
  mutate(EqIDNum = row_number())
equation <- equation %>%
  mutate(EqIDStr = "EQ")
equation <- equation %>%
  unite(EqID, EqIDNum, EqIDStr, sep="")

#Water Temperature
water_temp <- water_temp %>%
  mutate(WaterTempIDNum = row_number())
water_temp <- water_temp %>%
  mutate(WaterTempIDStr = "WTP")
water_temp <- water_temp %>%
  unite(WaterTempID, WaterTempIDNum, WaterTempIDStr, sep="")

water_temp <- water_temp %>% 
  mutate(TempUnits = "C")

#Swim Exhaustion Information
swim_ex_info <- swim_ex_info %>%
  mutate(SwimExIDNum = row_number())
swim_ex_info <- swim_ex_info %>%
  mutate(SwimExIDStr = "EXH")
swim_ex_info <- swim_ex_info %>%
  unite(SwimExID, SwimExIDNum, SwimExIDStr, sep="")

swim_ex_info <- swim_ex_info %>%
  mutate(TimeUnits = "s")

swim_ex_info <- swim_ex_info %>%
  mutate(VelocityUnits = "cm/s")

#Swim Variability
swim_var <- swim_var %>%
  mutate(SwimVarIDNum = row_number())
swim_var <- swim_var %>%
  mutate(SwimVarIDStr = "SVR")
swim_var <- swim_var %>%
  unite(SwimVarID, SwimVarIDNum, SwimVarIDStr, sep="")

swim_var <- swim_var %>%
  mutate(SwimVariabilityUnits = "cm/s")

#Fish Length Information
fish_len_info <- fish_len_info %>%
  mutate(FishIDNum = row_number())
fish_len_info <- fish_len_info %>%
  mutate(FishIDStr = "FSH")
fish_len_info <- fish_len_info %>%
  unite(FishID, FishIDNum, FishIDStr, sep="")

fish_len_info <- fish_len_info %>%
  mutate(FishVarIDNum = row_number())
fish_len_info <- fish_len_info %>%
  mutate(FishVarIDStr = "FVR")
fish_len_info <- fish_len_info %>%
  unite(FishVarID, FishVarIDNum, FishVarIDStr, sep="")

fish_len_info <- fish_len_info %>%
  mutate(LengthCentrTendMetric = "mean")

#Fish Length Variability
fish_len_var <- fish_len_var %>%
  mutate(FishVarIDNum = row_number())
fish_len_var <- fish_len_var %>%
  mutate(FishVarIDStr = "FVR")
fish_len_var <- fish_len_var %>%
  unite(FishVarID, FishVarIDNum, FishVarIDStr, sep="")

fish_len_var <- fish_len_var %>%
  mutate(LengthVariabilityUnits = LengthUnits)

#Experimental Method
exp_method <- exp_method %>%
  mutate(ExpMethodIDNum = row_number())
exp_method <- exp_method %>%
  mutate(ExpMethodIDStr = "EXP")
exp_method <- exp_method %>%
  unite(ExpMethodID, ExpMethodIDNum, ExpMethodIDStr, sep="")

```

**Next, the columns in each table were rearranged to match the order shown in the relational database model: **

```{r, warning=FALSE}
#Fish Information
fish_info <- fish_info %>%
  relocate(FishIDX, FishID, SwimID, ExpMethodID)

#Swimming Information
swim_info <- swim_info %>%
  relocate(SwimID, SwimExID, WaterTempID, SwimVarID, EqID)
swim_info <- swim_info %>%
  relocate(SwimCentrTendMetric, .after = SwimMode)
swim_info <- swim_info %>%
  relocate(TimeUnits, .after = SwimTime)

#Equation
equation <- equation %>%
  relocate(EqID)

#Water Temperature
water_temp <- water_temp %>%
  relocate(WaterTempID)

#Swim Exhaustion Information
swim_ex_info <- swim_ex_info %>%
  relocate(SwimExID)
swim_ex_info <- swim_ex_info %>%
  relocate(TimeUnits, .after = TimeBetweenIncrements)

#Swimming Variability
swim_var <- swim_var %>%
  relocate(SwimVarID)

#Fish Length Information
fish_len_info <- fish_len_info %>%
  relocate(FishID, FishVarID, LengthCentrTendMetric)

#Fish Length Variability
fish_len_var <- fish_len_var %>%
  relocate(FishVarID)

#Experimental Method
exp_method <- exp_method %>%
  relocate(ExpMethodID)
```

# Table Export:

**Once the tables were created, they were exported as .csv files to be later read in to SQL (see 04_product folder to view the tables): **

```{r}
write.csv(fish_info, "./01_acquisition/04_product/fish_info.csv", row.names = F)
write.csv(swim_info, "./01_acquisition/04_product/swim_info.csv", row.names = F)
write.csv(equation, "./01_acquisition/04_product/equation.csv", row.names = F)
write.csv(water_temp, "./01_acquisition/04_product/water_temp.csv", row.names = F)
write.csv(swim_ex_info, "./01_acquisition/04_product/swim_ex_info.csv", row.names = F)
write.csv(swim_var, "./01_acquisition/04_product/swim_var.csv", row.names = F)
write.csv(fish_len_info, "./01_acquisition/04_product/fish_len_info.csv", row.names = F)
write.csv(fish_len_var, "./01_acquisition/04_product/fish_len_var.csv", row.names = F)
write.csv(exp_method, "./01_acquisition/04_product/exp_method.csv", row.names = F)
```











